---
title: "Reproducible Research: Peer Assessment 1"
author: "Caitong YE"
date: "16 April 2019"
output: html_document
keep_md: true
---
## load packages
```{r, echo=TRUE}
library(lubridate)
library(dplyr)
library(ggplot2)
library(lattice)
```

## Loading and preprocessing the data
```{r, echo = TRUE}
# set work directory
# setwd("Repository/coursera/05_reproducible_research/RepData_PeerAssessment1/")
# unzip data file 
# unzip("activity.zip")

# load data
activity <- read.csv("activity.csv")

# general info
str(activity)

# format data
activity$date <- as.Date(activity$date)

```
## What is mean total number of steps taken per day?
```{r, echo=TRUE}
# 1 - Total steps per day
steps_per_day <- aggregate(steps ~ date, activity, sum)

# 2 - Histogram
png(filename="1_Histogram1")
hist(steps_per_day$steps, breaks=10, col="blue", main="Total Steps Per day", 
     xlab="Steps", ylab = "Frequency"  )
dev.off()

# 3 - Mean and median steps 
steps_mean <- mean(steps_per_day$steps)
steps_median <- median(steps_per_day$steps)
steps_mean
steps_median
```

- The Mean = 10766.19, and the Median = 10765

## What is the average daily activity pattern?
```{r, echo=TRUE}
# 1 - Average steps by interval
steps_by_interval <- aggregate(steps ~ interval, activity, mean)

# 2 - Time series plot
png(filename="2_TimeSeries")
plot(steps_by_interval$interval, steps_by_interval$steps, type="l", 
     xlab="Interval", ylab="Number of steps", 
     main="Average number of steps per day by interval")
dev.off()

# 3 - Max interval
max_interval <- steps_by_interval[which.max(steps_by_interval$steps), ]
max_interval
```

- 5-minute interval contains the maximum number of steps = 835
  with averaged steps = 206

## Imputing missing values
```{r, echo=TRUE}
# 1 - Total NAs
NAs <- sum(is.na(activity$steps))
NAs

NAs_percent <- mean(is.na(activity$steps))
print(paste0("Percentage of missing values: ", round(NAs_percent*100, 1), "%"))

# 2 - Impute by mean for that 5-mins interval 
imputed_activity <- activity
imputed_activity <- mutate(imputed_activity, steps=ifelse(is.na(steps),steps_by_interval$steps[match(interval, steps_by_interval$interval)], steps))
sum(is.na(imputed_activity))

# 3 - New histogram
new_steps_per_day <- aggregate(steps ~ date, imputed_activity, sum)
png(filename="3_Histogram2")
hist(new_steps_per_day$steps, breaks=10, col="blue", xlab="Steps", ylab="Frequency", 
     main="Total Steps Per Day after imputation")
dev.off()
mean_imputed <- mean(new_steps_per_day$steps)
median_imputed <- median(new_steps_per_day$steps)
mean_imputed
median_imputed

```
- The Mean after imputation = 10766.19, and the Median after imputation = 10766.19
- As we can see here, the Mean doesn't change after imputation due to the fact 
  that we impute the missing values with the average of 5-minutes of all days, 
  which results in an increase in total steps per day
- The Median shifted because there are now more values in the data

## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE}
# 1 - Add factor <wdays> to indicate weekdays and weekends
imputed_activity <- mutate(imputed_activity, wdays=ifelse(weekdays(date)=="Saturday"|weekdays(date)=="Sunday", "weekend", "weekday"))

# 2 - Calculate average steps per 5-mins interval 
imputed_activity <- aggregate(steps ~ interval + wdays, imputed_activity, mean)

# 3 - Panel plot 
png(filename="4_PanelPlot")
xyplot(steps ~ interval | wdays, imputed_activity, type="l", layout=(c(1, 2)), xlab="Interval", ylab="number of steps")
dev.off()
```
